Reformulating phylogenetic mixed models to improve flexibility and speed

### 2021-07-15 response.md

Thank you for submitting your research to MEE and apologies for the long time getting back to you, which partly a consequence of the ongoing lock-down work-from-home situation. Your manuscript has been reviewed by two highly qualified reviewers who found your work interesting while raising several points about the presentation and extent of the simulations, the scalability of the method and suggesting the inclusion of empirical examples. Based on the reviewers’ judgment and my own take on the paper I cannot recommend the acceptance of this paper at this point, but would encourage you to resubmit a thoroughly re-worked version.

**We appreciate the opportunity to resubmit a revised version.**

As currently presented I found this paper a little bit in between an application note and a research paper but lacking some key components from both, but turning it into a research paper would require, in my opinion, the addition of an empirical example (see also comments from reviewer 1).

**We intended the paper as an application/software paper, not as a research paper: there is some novelty here, but as our primary innovation is to recast the model in a mathematically equivalent but more convenient/efficient form, we would prefer to treat this as an application. We have provided an example in the vignette.**

@@

In the Introduction, a broader perspective about PCMs should be given before jumping into the specifics of the models tested here. Phylogenetic comparative methods include lots of other things, see e.g. Luke Harmon’s book: https://lukejharmon.github.io/pcm/chapters/. In general, a better reference to previous work is required and some claims of the paper should be either demonstrated or removed. For example, stating that “existing [PCM] procedures are either insufficiently flexible or too computationally demanding to analyze large data sets” is simply not true. There are many research papers applying PCMs to large datasets (and as a side note: how large should a dataset be to be considered large?).

**Todo: Revisit and rewrite. It is true many research papers used PCMs to large datasets, but it depends on the complexity of the model. I.e. If we simply use GLM methods in Table 1, it is relative fast even with large datasets. If we have more complicated models, then it will be very computationally demanding.**

A sentence like “Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely” needs references supporting it. For example, Kostikova et al. 2016 (DOI:10.1093/sysbio/syw010) present a model that seems relevant here. Further, claiming that ‘brms’ is more efficient than ‘MCMCglmm’ needs to be backed up by either a reference or a test (see also comments from the reviewers).

** This line is now removed. We have confirmed that the suggested example is relevant to our framework in partitioning different levels of variation and cited appropriately. **

@@

In the methods it should be better clarified what is the state of the art and what are the novel parts you are presenting in this study.

**Bayesian methods are the state of the art for more complex. I've now reworded the last paragraph in the intro before methods.**

@@

In lines 74-75 “phylogenetic mixed models in community ecology” are mentioned but it is unclear what the link is to the previous paragraph and the to the following sentences.

**This refers to the recent developed frequentist tools for phyloglmm and used in community ecology applications. We have now reword and clarify this sentence. **

@@

Why are the parameter estimates of the two implementations of the new model different in Fig. 5? And, in the same figure, why were runs of alternative implementations killed after 30 min?

**MLi: The parameters that are in both single and multiple site models remain the same. We have now added one more alternative implementation run for the 100 species simulation with the slow platforms. The goal of this paper is to show the inflexiably of existing tools and the efficiency of the new implemented approach. The implementation of glmmTMB designed to maximize speed and flexibility. All else equal, the computational performance of glmmTMB are expected to outperform lme4. We added a text in the results that cites this expectiation.**

@@

**MLi: Thanks.**

However, I think the article could fail to reach the extent of its audience by missing several key elements:

- Because that work isn’t a theoretical work but instead the description of an applied statistics method, adding a few examples, such as the one provided at l68-73, would help the reader understanding the value of those models. I think examples should particularly be aimed at the “Simulation” § l132-155, where they could shed light on the subtle differences between the parameters

**MLi: We have added an example in the vignette for this package**

@@

- In the same spirit of applied statistics, a brief description (or table) of which functions to use in the libraries lme4 and glmmTMB to make these models work would make it easier to start. There should also be mention of the tutorial available in the repository

**MLi: We have added an example in the vignette for this package**

- It would also be beneficial to add an application of the method with biological data directly in the article. The more your article is associated with biological ideas, the more users you will reach in my opinion.

**MLi: Thanks for the suggestion. This paper is written to be an application paper rather than a research paper. We have provided a real example from Garamszegi's ``Modern phylogenetic comparative methods and their application in evolutionary biology: concepts and practice'' in the repository.**

@@

- Naming that class of models (even though it is still pglmm) could help the reader follow the results section and identify the method as well.

**Thanks. We have now renamed everything under phylogenetic regression**

@@

These remarks might sound cosmetic but I think they might help to reach a broader audience and help all kind of users getting started with those models.

I recommend the authors to think on those ideas to improve the visibility of their method and the journal to ask them to do so in order to increase the impact of the article.

In a purely scientific perspective, I have no objections to publish that article.

**Thanks, these are wonderful and helpful suggestion. We have included a tutorial example in the released version on zenodo/github. We have now included the link in the paper.... **

### 2021-07-15 phyloglmm_ms_resubmit.tex
pushing new changes in the response and added glmmTMB citation

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/msplot.pdf}
  \caption{Comparison of multi-group model parameter estimates. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters. \pkg{pez} and \pkg{phyr} estimates for n = 100 and 500 are missing because the models did not converge within 30 minutes.
  }
  \label{msplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mstime.pdf}
  \caption{Comparison of multi-group model computational speed.}
  \label{msplot_time}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
  \caption{Comparison of multi-group model coverage. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
  \label{msplot_coverage}
\end{figure}
\end{center}

@@

Although the parameter estimates are similar across platforms, computational efficiency varies enormously across platforms and sample size.
Comparing our two implementations, \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}.
\pkg{glmmTMB} is also faster than \pkg{pez} and \pkg{phyr}: the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9, versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr}.
\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was impractical to fit 500-species models with \pkg{pez} and \pkg{phyr}.

>>>>

\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/msplot.pdf}
  \caption{Comparison of multi-group model parameter estimates. The horizontal line shows the true value of the parameters in the simulation model. Models capable of fitting all parameters (\pkg{lme4} and \pkg{glmmTMB}) fit well for all parameters. \pkg{pez} and \pkg{phyr} estimates for n = 500 are missing because the models did not converge within 30 minutes.
  }
  \label{msplot}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mstime.pdf}
  \caption{Comparison of multi-group model computational speed.}
  \label{msplot_time}
\end{figure}
\end{center}
\begin{center}
\begin{figure}[H]
  \includegraphics[scale=0.7]{./figure/mscoverage.pdf}
  \caption{Comparison of multi-group model coverage. The black line shows the nominal coverage, and the blue ribbon the 95\% binomial confidence interval based on 100 simulated fits.}
  \label{msplot_coverage}
\end{figure}
\end{center}

@@

Although the parameter estimates are similar across platforms, computational efficiency varies enormously across platforms and sample size.
Comparing our two implementations, \pkg{glmmTMB} is almost an order of magnitude faster than \pkg{lme4}; \pkg{glmmTMB} is expected to out perform \pkg{lme4} as it is designed to maximize flexibility and speed \citep{brooks2017glmmTMB}.
\pkg{glmmTMB} is also faster than \pkg{pez} and \pkg{phyr}: the median time for \pkg{glmmTMB} to fit a 50-species model is $\approx$ 9, versus $\approx$ 200 seconds for \pkg{pez} and \pkg{phyr}.
\pkg{glmmTMB} takes $\approx$ 125 seconds to fit a 500-species model; it was impractical to fit 500-species models with \pkg{pez} and \pkg{phyr}.
